---
title: "Data_analysis"
author: "Shayle Matsuda"
date: "2025-12-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Load Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

library(lmerTest)
library(car)
library(emmeans)

library(vegan) 
library(pairwiseAdonis)

library(phyloseq)
library(plyr)

library(viridis)
library(lme4)
library(lmerTest)

library(glmmTMB)
library(ANCOMBC)


library(microbiome)  
library(tidyverse)
```

Load data and Clean up
```{r}
load("RData/Bac.3k.RData")
Bac.seq

Bac.seq<-subset_samples(Bac.seq, Type=="sample") #filter out only samples
sd <- as(sample_data(Bac.seq), "data.frame") #look at samp data, n=245

Bac.seq.ACER<-subset_samples(Bac.seq, species.x=="Acer")
sd.ACER <- as(sample_data(Bac.seq.ACER), "data.frame") #look at samp data

Bac.seq.ACER<-subset_samples(Bac.seq.ACER, Date!="5/16/23") #remove antibac T0 samps
sd.ACER <- as(sample_data(Bac.seq.ACER), "data.frame") #look at samp data, n=206


#Make rel abund df
Bac.seq.rel  = transform_sample_counts(Bac.seq.ACER, function(x) x / sum(x) )  #save as RELA DIVs
Bac.seq.rel.sd <- as(sample_data(Bac.seq.rel), "data.frame") #look at samp data
Bac.seq.rel = prune_taxa(taxa_sums(Bac.seq.rel) > 0, Bac.seq.rel) #this removes any OTU with 0s
Bac.seq.rel.sd$Tube<-as.numeric(Bac.seq.rel.sd$Tube)
#number of samples/treat/time
num<-Bac.seq.rel.sd %>%
  dplyr::count(tank.x, Date) %>%
  arrange(tank.x, Date)
#write.csv(num, "acer samp per treat time.csv")
```

Bar plots
```{r}
df <- phyloseq::psmelt(Bac.seq.rel)


df$Date <- as.Date(as.character(df$Date), format = "%m/%d/%y")
df <- df %>%
  dplyr::arrange(Date) %>%
  dplyr::mutate(Date = factor(Date, levels = unique(Date)))

df_avg <- df %>%
  dplyr::group_by(Geno, tank.x, Date, Genus) %>%
  dplyr::summarise(Abundance = mean(Abundance), .groups = "drop")




plots_by_geno <- df_avg %>%
  split(.$Geno) %>%
  imap(function(Date, geno_name) {
    ggplot(Date, aes(x = Date, y = Abundance, fill = Genus)) +
      geom_bar(stat = "identity") +
      facet_wrap(~ tank.x, scales = "free_x") +
      theme_bw() +
      theme(
        legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
      ) +
      ylab("Relative Abundance") +
      xlab("Date") +
      ggtitle(paste("Genotype:", geno_name))
  });plots_by_geno

# Create a folder to store PDFs (optional)
dir.create("plots_family", showWarnings = FALSE)

# Loop and save each plot using genotype name + "Family"
for(geno_name in names(plots_by_geno)) {
  
  pdf(file = paste0("plots_family/", geno_name, "_Genus.pdf"),
      width = 8, height = 6)   # adjust size if needed
  
  print(plots_by_geno[[geno_name]])
  
  dev.off()
}

# plots by tank ####

# Agglomerate to family (change taxrank if needed)
ps_fam <- tax_glom(Bac.seq.rel, taxrank = "Family")


# Melt to long data frame
df_plot <- psmelt(ps_fam)

# Make Date an actual date if it isn’t
df_plot$Date <- as.Date(df_plot$Date, format = "%m/%d/%y")  # adjust format if needed

# Sort Dates if you want (optional)
df_plot <- df_plot %>% arrange(Date)

# Plot
ggplot(df_plot, aes(x = Date, y = Abundance, fill = Family)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ tank.x, scales = "free_x") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    x = "Date",
    y = "Relative Abundance",
    title = "Community Composition by Date, Faceted by Tank"
  )

library(dplyr)

df_plot_sum <- df_plot %>%
  group_by(tank.x, Date, Family) %>%        # <-- tank.x included here
  dplyr::summarise(
    Abundance = mean(Abundance),
    .groups = "drop"
  )

ggplot(df_plot_sum, aes(x = Date, y = Abundance, fill = Family)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ tank.x, scales = "free_x") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    x = "Date",
    y = "Relative Abundance",
    title = "Community Composition by Date, Faceted by Tank"
  )

```

Top 10 otus: Phyla, Fam, Genus
```{r}
#
# Selecting top 10 OTUs
TopNOTUs = names(sort(taxa_sums(Bac.seq.rel), TRUE)[1:10])
acer.abund = prune_taxa(TopNOTUs, Bac.seq.rel)
acer.abund.melt <- psmelt(acer.abund)
acer.abund.melt$top.group <- "abund"
acer.abund.melt$Date <- as.Date(acer.abund.melt$Date, format = "%m/%d/%y")
acer.abund.melt <- acer.abund.melt %>%
  arrange(Date)
#write.csv(acer.abund.melt, "outputs/acer_abund.csv")

acer.list <- unique(acer.abund.melt$Genus) #tops

#Genus

sum <- ddply(acer.abund.melt, c( "tank.x", "Date", "Genus"), summarise,
             N = length(Abundance), 
             mean = mean(Abundance),
             sd = sd(Abundance), 
             se = sd/sqrt(N)
)

pal <- c(
  "#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD",
  "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17BECF",
  "#393B79", "#637939", "#8C6D31", "#843C39", "#7B4173",
  "#A55194", "#396AB1", "#DA7C30", "#3E9651", "#CC2529"
)

ggplot(sum, aes(x = Date, y = mean, group = Genus, color = Genus)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                alpha = 0.5, width = 0) +
  geom_line(alpha = 0.5) +
  scale_color_manual(values = pal) +     # <--- use color, not fill
  ylab("Mean Relative Abundance\n") +
  xlab("\nDate") +                       # x is Date here, not Treatment
  labs(color = "OTU Genus") +
  theme_bw() +
  facet_wrap(~ tank.x)

#ggsave(filename = "outputs/acer_treatment_by_abund.pdf", plot = last_plot())

#Phylum
sum <- ddply(acer.abund.melt, c( "tank.x", "Date", "Phylum"), summarise,
             N = length(Abundance), 
             mean = mean(Abundance),
             sd = sd(Abundance), 
             se = sd/sqrt(N)
)

ggplot(sum, aes(x = Date, y = mean, group = Phylum, color = Phylum)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                alpha = 0.5, width = 0) +
  geom_line(alpha = 0.5) +
  scale_color_manual(values = pal) +     # <--- use color, not fill
  ylab("Mean Relative Abundance\n") +
  xlab("\nDate") +                       # x is Date here, not Treatment
  labs(color = "OTU Phylum") +
  theme_bw() +
  facet_wrap(~ tank.x)

#ggsave(filename = "outputs/acer_treatment_by_abund_phy.pdf", plot = last_plot())


#Family
sum <- ddply(acer.abund.melt, c( "tank.x", "Date", "Family"), summarise,
             N = length(Abundance), 
             mean = mean(Abundance),
             sd = sd(Abundance), 
             se = sd/sqrt(N)
)

ggplot(sum, aes(x = Date, y = mean, group = Family, color = Family)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                alpha = 0.5, width = 0) +
  geom_line(alpha = 0.5) +
  scale_color_manual(values = pal) +     # <--- use color, not fill
  ylab("Mean Relative Abundance\n") +
  xlab("\nDate") +                       # x is Date here, not Treatment
  labs(color = "OTU Family") +
  theme_bw() +
  facet_wrap(~ tank.x)

#ggsave(filename = "outputs/acer_treatment_by_abund_fam.pdf", plot = last_plot())


########### Remake plots without MD3-55 to see trends
sum.no.MD3 <- sum %>% 
  filter(Genus != "MD3-55")

ggplot(sum.no.MD3, aes(x = Date, y = mean, group = Genus, color = Genus)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                alpha = 0.6, width = 0) +
  geom_line(alpha = 0.7) +
  scale_color_manual(values = pal) +
  ylab("Mean Relative Abundance (excluding MD3-55)\n") +
  xlab("\nDate") +
  labs(color = "OTU Genus") +
  theme_bw() +
  facet_wrap(~ tank.x)

#ggsave(filename = "outputs/acer_treatment_by_abund_gen_noMD3.pdf", plot = last_plot())

#Fam
sum.no.MD3 <- sum %>% 
  filter(Family != "Midichloriaceae")

ggplot(sum.no.MD3, aes(x = Date, y = mean, group = Family, color = Family)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                alpha = 0.6, width = 0) +
  geom_line(alpha = 0.7) +
  scale_color_manual(values = pal) +
  ylab("Mean Relative Abundance (excluding MD3-55)\n") +
  xlab("\nDate") +
  labs(color = "OTU Family") +
  theme_bw() +
  facet_wrap(~ tank.x)

#ggsave(filename = "outputs/acer_treatment_by_abund_fam_noMD3.pdf", plot = last_plot())

```

PERMANOVA All times
```{r}
#make unifrac matrix
Bac.seq.rel.wu <- phyloseq::distance(Bac.seq.rel, method = "wunifrac")
Bac.seq.rel.wu.df <- as(sample_data(Bac.seq.rel), "data.frame") #sample df

#adonis
set.seed(30) #~Date
adonis2(Bac.seq.rel.wu ~ Date, data = Bac.seq.rel.wu.df) # p=0.04

set.seed(30)  #~Tank Treatment
adonis2(Bac.seq.rel.wu ~ tank.x, data = Bac.seq.rel.wu.df) # p=0.017

set.seed(30)  #~Genotype
adonis2(Bac.seq.rel.wu ~ Geno, data = Bac.seq.rel.wu.df) # p=0.001

#1 Does tank explain microbiome differences overall? Significant
#Ignoring Date, controling for genotype
set.seed(30)  #~Genotype + Tank, under a reduced model.  
adonis2(Bac.seq.rel.wu ~ Geno + tank.x,data = Bac.seq.rel.wu.df,by = "margin")

  #adding in strata to account for ramets of each genet in each tank
  adonis2(
    Bac.seq.rel.wu ~ tank.x,
    data = Bac.seq.rel.wu.df,
    by = "margin",
    strata = Bac.seq.rel.wu.df$Geno
  )

#2 Do tanks follow different temporal trajectories? Do microbiomes change through time differently in different tanks? Not Significant
set.seed(30)
adonis2(
  Bac.seq.rel.wu ~ Geno + Date * tank.x,
  data = Bac.seq.rel.wu.df,
  by = "margin"
)

# Pairwise PERMANOVA between tanks, controlling for genotype ####
# this helps identify which tanks are driving the differences

tanks <- unique(Bac.seq.rel.wu.df$tank.x)

pairwise_res <- combn(tanks, 2, simplify = FALSE, FUN = function(x) {
  df_sub <- Bac.seq.rel.wu.df %>% filter(tank.x %in% x)

  dist_sub <- as.dist(as.matrix(Bac.seq.rel.wu)[rownames(df_sub), rownames(df_sub)])

  ad <- adonis2(
    dist_sub ~ Geno + tank.x,
    data = df_sub,
    by = "margin",
    permutations = 999
  )

  data.frame(
    tank1 = x[1],
    tank2 = x[2],
    R2 = ad["tank.x", "R2"],
    p = ad["tank.x", "Pr(>F)"]
  )
})

pairwise_df <- bind_rows(pairwise_res)

# FDR correction
pairwise_df$p_adj <- p.adjust(pairwise_df$p, method = "fdr")

pairwise_df %>% arrange(p_adj) #of the 5 pairs that are sig, NWQ4 is part of 4. 


```
PERMANOVA - Looking at the question - Microbiomes adjust rapidly after exposure to a new species context, then stabilize.
Here, we test Time 0 (4/3) to all later time points -Is the microbiome at the initial state different from the post-adjustment state?
```{r}
Bac.seq.rel.wu.df <- Bac.seq.rel.wu.df %>%
  dplyr::mutate(
    Date = as.Date(Date, format = "%m/%d/%y")
  )
#Create a binary “pre vs post” time variable
Bac.seq.rel.wu.df <- Bac.seq.rel.wu.df %>%
  dplyr::mutate(
    TimePhase = ifelse(Date == as.Date("2023-04-03"),
                       "T0",
                       "Post")
  )
Bac.seq.rel.wu.df <- Bac.seq.rel.wu.df %>% #order
  dplyr::mutate(
    TimePhase = factor(TimePhase, levels = c("T0", "Post"))
  )

#permanova
set.seed(30)
adonis2(
  Bac.seq.rel.wu ~ Geno + TimePhase,
  data = Bac.seq.rel.wu.df,
  by = "margin"
) #not significant p=0.09

#dispersion - not significant p= 0.14
bd_time <- betadisper(Bac.seq.rel.wu, Bac.seq.rel.wu.df$TimePhase)
anova(bd_time)



```


testing within each tank/geno, prob dont need due to low n
```{r}
# test Date within each tank - p>0.05 so within tanks nothing changes over time ####
# make sure rownames are right
stopifnot(all(attr(Bac.seq.rel.wu, "Labels") == rownames(Bac.seq.rel.wu.df)))

tank_idx_list <- split(seq_len(nrow(Bac.seq.rel.wu.df)), Bac.seq.rel.wu.df$tank.x)

res_by_tank <- lapply(tank_idx_list, function(idx) {
  # subset distance matrix to those samples
  dist_sub <- as.dist(as.matrix(Bac.seq.rel.wu)[idx, idx])

  # subset metadata
  meta_sub <- Bac.seq.rel.wu.df[idx, , drop = FALSE]

  adonis2(dist_sub ~ Date, data = meta_sub)
})

res_by_tank

# test Date within each Geno - Genotype Y27 is the only one p=0.001, 

geno_idx_list <- split(seq_len(nrow(Bac.seq.rel.wu.df)), Bac.seq.rel.wu.df$Geno)

res_by_geno <- lapply(geno_idx_list, function(idx) {
  dist_sub <- as.dist(as.matrix(Bac.seq.rel.wu)[idx, idx])
  meta_sub <- Bac.seq.rel.wu.df[idx, , drop = FALSE]
  adonis2(dist_sub ~ Date, data = meta_sub)
})

res_by_geno

##### Ok let's try again with date as numeric ####
Bac.seq.rel.wu.df$Date <- as.Date(Bac.seq.rel.wu.df$Date, format="%m/%d/%y")

Bac.seq.rel.wu.df$DayNum <- as.numeric(Bac.seq.rel.wu.df$Date) -
                             min(as.numeric(Bac.seq.rel.wu.df$Date))

#1) db-RDA (distance-based redundancy analysis) Uses genotype/tank as covariates, time as numeric: Is there a directional pattern in community composition that can be explained by a predictor, after accounting for other sources of variation? is there a directional gradient?

cap <- capscale(Bac.seq.rel.wu ~ DayNum + Condition(Geno) + Condition(tank.x),
                data = Bac.seq.rel.wu.df)
anova(cap) # p=0.014


#2) NMDS axis regression 
library(vegan)

# 1. Run the NMDS
nmds <- metaMDS(Bac.seq.rel.wu)

# 2. Get site scores explicitly from vegan, and don't call the object 'scores'
site.scores <- vegan::scores(nmds, display = "sites")

# 3. Put into a data.frame with your DayNum
df_nmds <- data.frame(
  MDS1   = site.scores[, 1],
  MDS2   = site.scores[, 2],
  DayNum = Bac.seq.rel.wu.df$DayNum
)

# 4. Regress MDS1 on time
m1 <- lm(MDS1 ~ DayNum, data = df_nmds)
anova(m1) # p=0.0133
summary(m1) # see notes


## Do this for JUST the significant genotype Y27
idx_Y27 <- Bac.seq.rel.wu.df$Geno == "Y27"

dist_Y27 <- as.dist(as.matrix(Bac.seq.rel.wu)[idx_Y27, idx_Y27])
meta_Y27 <- Bac.seq.rel.wu.df[idx_Y27, , drop = FALSE]

nmds_Y27 <- metaMDS(dist_Y27)
site.scores_Y27 <- vegan::scores(nmds_Y27, display = "sites")

df_Y27 <- data.frame(
  MDS1   = site.scores_Y27[,1],
  MDS2   = site.scores_Y27[,2],
  DayNum = meta_Y27$DayNum
)

m_Y27 <- lm(MDS1 ~ DayNum, data = df_Y27)
anova(m_Y27)
summary(m_Y27)


m_Y27 <- lm(MDS1 ~ DayNum, data = df_Y27)
anova(m_Y27)
summary(m_Y27)


```

# Homogeneity of dispersion test  -
```{r}
bd <- betadisper(Bac.seq.rel.wu, Bac.seq.rel.wu.df$Date)
anova(bd)            # significance test
TukeyHSD(bd)         # which dates differ


# index for Y27 only
idx_Y27 <- Bac.seq.rel.wu.df$Geno == "Y27"

dist_Y27 <- as.dist(as.matrix(Bac.seq.rel.wu)[idx_Y27, idx_Y27])
meta_Y27 <- Bac.seq.rel.wu.df[idx_Y27, , drop = FALSE]

# betadisper for Y27 only NON SIG
bd_Y27 <- betadisper(dist_Y27, meta_Y27$Date)
anova(bd_Y27)   #p >0.05
TukeyHSD(bd_Y27)

plot(bd_Y27, hull=FALSE)
boxplot(bd_Y27)

```


Delete?
Question 1 - more explicitly - Do microbiomes change over time within each treatment tank?
time × treatment
```{r}
#Test time within each tank, accounting for genotype - Tab 2 Q1 Adonis
# distance ~ Geno + Date   (Type II / marginal)in 
# Do microbiomes change over time within each treatment tank, accounting for genotype?

res_by_tank <- lapply(
  split(seq_len(nrow(Bac.seq.rel.wu.df)), Bac.seq.rel.wu.df$tank.x),
  function(idx) {
    adonis2(
      as.dist(as.matrix(Bac.seq.rel.wu)[idx, idx]) ~
        Geno + Date,
      data = Bac.seq.rel.wu.df[idx, ],
      by = "margin"
    )
  }
)



```

Rerun these:
Main inferential analysis for replicated treatments only (Control vs Mixed ONLY). 
Geno explains 33% of variation. There is no difference between treatments across time. 
Treatment:Date p=0.10
Treatment:Time p=0.23
```{r}
#1) Make a Treatment factor and keep TankID
df_cm <- subset(Bac.seq.rel.wu.df, tank.x %in% c("FL_Control_1","FL_Control_2","Mixed_1","Mixed_2"))

df_cm$Treatment <- ifelse(grepl("^FL_Control", df_cm$tank.x), "Control", "Mixed")
df_cm$Treatment <- factor(df_cm$Treatment)
df_cm$TankID <- factor(df_cm$tank.x)

dist_cm <- as.dist(as.matrix(Bac.seq.rel.wu)[rownames(df_cm), rownames(df_cm)])

#2) Test whether treatments change differently through time, accounting for genotype and tank within treatment: Not Significant, 

perm_tank <- how(blocks = df_cm$TankID)
set.seed(30)

adonis2(
  dist_cm ~ Geno + Treatment*Date + Treatment:TankID,
  data = df_cm,
  by = "margin",
  permutations = perm_tank
)

#testing differences only at endpoint: Not significant

df_end <- subset(df_cm, Date == max(Date))
dist_end <- as.dist(as.matrix(dist_cm)[rownames(df_end), rownames(df_end)])

perm_end <- how(blocks = df_end$TankID)
set.seed(1)

adonis2(
  dist_end ~ Geno + Treatment + Treatment:TankID,
  data = df_end,
  by = "margin",
  permutations = perm_end
)

#Conclusion: The mixed treatment does not produce a consistent microbiome response across tanks (Mixed_2 is the only one significant)

```

#NMDS plots
```{r}

# NMDS on the full distance matrix
set.seed(1)
nmds <- metaMDS(Bac.seq.rel.wu, k = 2, trymax = 50)

# Extract site scores
site <- as.data.frame(vegan::scores(nmds, display = "sites"))
site$SampleID <- rownames(site)

# Join metadata
df_plot <- Bac.seq.rel.wu.df %>%
  mutate(SampleID = rownames(Bac.seq.rel.wu.df)) %>%
  inner_join(site, by = "SampleID")

# Make sure Date is an actual Date and ordered
# Use the format that matches your data; %m/%d/%y worked for you earlier
if (!inherits(df_plot$Date, "Date")) {
  df_plot$Date <- as.Date(df_plot$Date, format = "%m/%d/%y")
}
df_plot <- df_plot %>% arrange(tank.x, Geno, Date)

df_plot <- df_plot %>%
  dplyr::mutate(
    Date = as.Date(Date),
    Date_f = factor(Date, levels = sort(unique(Date)))
  )

n_dates <- nlevels(df_plot$Date_f)
pal_date <- grDevices::rainbow(n_dates)

ggplot(df_plot, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = Date_f), size = 2, alpha = 0.9) +
  facet_wrap(~ tank.x) +
  scale_color_manual(values = pal_date) +
  theme_bw() +
  labs(
    title = "NMDS of microbiome composition",
    subtitle = "Faceted by tank; points colored by sampling date",
    color = "Date"
  )

#now add convex hulls around dates
hulls_date <- df_plot %>%
  dplyr::group_by(tank.x, Date_f) %>%
  dplyr::filter(n() >= 3) %>%                     # need ≥3 points for a hull
  slice(chull(NMDS1, NMDS2)) %>%
  ungroup()

ggplot(df_plot, aes(x = NMDS1, y = NMDS2)) +
  # date polygons
  geom_polygon(
    data = hulls_date,
    aes(fill = Date_f, group = interaction(tank.x, Date_f)),
    alpha = 0.2,
    color = NA
  ) +
  # points
  geom_point(
    aes(color = Date_f),
    size = 2,
    alpha = 0.9
  ) +
  facet_wrap(~ tank.x) +
  scale_color_manual(values = pal_date) +
  scale_fill_manual(values = pal_date) +
  theme_bw() +
  labs(
    title = "NMDS of microbiome composition",
    subtitle = "Convex hulls indicate sampling dates; faceted by tank",
    color = "Date",
    fill  = "Date"
  )

#ggsave(filename = "outputs/acer_nmds_bytankRainbow.pdf", plot = last_plot())

 
# facet by Date ####
n_tanks <- length(unique(df_plot$tank.x))
tank_pal <- hcl.colors(n_tanks, palette = "Dark 3")

hulls_tank <- df_plot %>%
  dplyr::group_by(Date_f, tank.x) %>%
  dplyr::filter(dplyr::n() >= 3) %>%   # need ≥3 points per hull
  dplyr::slice(chull(NMDS1, NMDS2)) %>%
  dplyr::ungroup()

n_tanks <- length(unique(df_plot$tank.x))
tank_pal <- hcl.colors(n_tanks, palette = "Dark 3")

ggplot(df_plot, aes(x = NMDS1, y = NMDS2)) +
  # tank polygons (within each Date facet)
  geom_polygon(
    data = hulls_tank,
    aes(
      fill = tank.x,
      group = interaction(Date_f, tank.x)
    ),
    alpha = 0.2,
    color = NA
  ) +
  # points
  geom_point(
    aes(color = tank.x),
    size = 2,
    alpha = 0.9
  ) +
  facet_wrap(~ Date_f) +
  scale_color_manual(values = tank_pal) +
  scale_fill_manual(values = tank_pal) +
  theme_bw() +
  labs(
    title = "NMDS of microbiome composition",
    subtitle = "Faceted by sampling date; colors and hulls indicate tank",
    color = "Tank",
    fill  = "Tank"
  )
#ggsave(filename = "outputs/acer_nmds_by_date.pdf", plot = last_plot())


#plot with Centroids per tank (No date)  ####
#NMDS - centroids to show tank differences
# ---- Highlight tanks ----
df_plot <- df_plot %>%
  dplyr::mutate(
    tank_highlight = dplyr::case_when(
      tank.x == "NWQ4" ~ "NWQ4",
      tank.x == "P4"   ~ "P4",
      TRUE             ~ "Other"
    )
  )

# ---- Compute centroids ----
tank_centroids <- df_plot %>%
  dplyr::group_by(tank.x) %>%
  dplyr::summarise(
    NMDS1 = mean(NMDS1, na.rm = TRUE),
    NMDS2 = mean(NMDS2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    tank_highlight = dplyr::case_when(
      tank.x == "NWQ4" ~ "NWQ4",
      tank.x == "P4"   ~ "P4",
      TRUE             ~ "Other"
    )
  )

p <- ggplot() +
  # background sample points
  geom_point(
    data = df_plot,
    aes(x = NMDS1, y = NMDS2, color = tank_highlight),
    size = 1.8,
    alpha = 0.5,
    position = position_jitter(width = 0.005, height = 0.005)
  ) +
  # centroid points
  geom_point(
    data = tank_centroids,
    aes(x = NMDS1, y = NMDS2, fill = tank_highlight),
    shape = 21,
    size = 5,
    color = "black",
    stroke = 0.6
  ) +
  # label NWQ4 centroid
  geom_text(
    data = subset(tank_centroids, tank.x == "NWQ4"),
    aes(x = NMDS1, y = NMDS2, label = "NWQ4"),
    vjust = -1.2,
    fontface = "bold",
    size = 4,
    inherit.aes = FALSE
  ) +
  # label P4 centroid
  geom_text(
    data = subset(tank_centroids, tank.x == "P4"),
    aes(x = NMDS1, y = NMDS2, label = "P4"),
    vjust = -1.2,
    fontface = "bold",
    size = 4,
    inherit.aes = FALSE
  ) +
  # colors
  scale_color_manual(
    values = c(
      "NWQ4"  = "#d73027",  # red
      "P4"    = "#4575b4",  # blue
      "Other" = "grey60"
    ),
    guide = "none"
  ) +
  scale_fill_manual(
    values = c(
      "NWQ4"  = "#d73027",
      "P4"    = "#4575b4",
      "Other" = "white"
    ),
    guide = "none"
  ) +
  coord_cartesian(
    xlim = c(-0.02, 0.06),
    ylim = c(-0.01, 0.01)
  ) +
  theme_bw() +
  labs(
    title = "NMDS of microbiome composition",
    subtitle = "Tank centroids shown; NWQ4 (red) and P4 (blue) highlighted"
  );p

#ggsave(filename = "outputs/acer_nmds_tankCentroids.pdf",plot = p) 


# Compare centroid distance to dispersion explicitly ####
#dispersion differs significantly between tanks
bd_tank <- betadisper(Bac.seq.rel.wu, Bac.seq.rel.wu.df$tank.x)
anova(bd_tank) #p=0.0022
TukeyHSD(bd_tank)

pairwise.adonis2(
  Bac.seq.rel.wu ~ tank.x,
  data = Bac.seq.rel.wu.df,
  strata = Bac.seq.rel.wu.df$Geno
)

#Next -Are NWQ4 and P4 different in within-tank variability (distance to centroid)?
disp_df <- data.frame(
  Sample   = rownames(Bac.seq.rel.wu.df),
  tank.x   = Bac.seq.rel.wu.df$tank.x,
  Distance = bd_tank$distances
)

disp_NWQ4_P4 <- disp_df %>% #subset to NWQ4 and P4
  dplyr::filter(tank.x %in% c("NWQ4", "P4"))

#Wilcoxon test
wilcox.test(Distance ~ tank.x, data = disp_NWQ4_P4) #W = 403, p-value = 0.06441, cannot say tank dispersion differs

#Is one tank more heterogeneous than the other? Welch t-test p = 0.876, mean(NWQ4) = 0.03435, mean(P4)= 0.03685, basically the same so dispersion like avg dispersion the same 
t.test(Distance ~ tank.x, data = disp_NWQ4_P4) 

#plot
geom_text(
  data = med_df,
  aes(
    x = tank.x,
    y = median_dist,
    label = sprintf("%.3f", median_dist)
  ),
  vjust = -1.2,
  fontface = "bold",
  size = 4)

ggplot(disp_NWQ4_P4, aes(x = tank.x, y = Distance, fill = tank.x)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("NWQ4" = "#d73027", "P4" = "grey70")) +
  geom_text(
  data = med_df,
  aes(
    x = tank.x,
    y =  median_dist + 0.002,
    label = sprintf("%.3f", median_dist)
  ),
  vjust = -1.2,
  fontface = "bold",
  size = 4)+
theme_bw() +
  labs(
    x = "Tank",
    y = "Distance to centroid",
    title = "Within-tank microbiome variability",
    subtitle = "NWQ4 vs P4"
  ) +
  theme(legend.position = "none")+annotate(
  "text",
  x = 1,
  y = max(disp_NWQ4_P4$Distance) * 0.95,
  label = "Medians do not differ (Wilcoxon p = 0.17)",
  size = 4,
  fontface = "italic"
)

#ggsave(filename = "outputs/acer_boxplot_inTankVariationP4NWQ4.pdf", plot = last_plot())


# plot distance to centroid for all tanks ####


# Build dispersion dataframe

# 1) Per-sample dispersion table from betadisper output
disp_all <- tibble(
  Sample   = rownames(Bac.seq.rel.wu.df),
  tank.x   = Bac.seq.rel.wu.df$tank.x,
  Distance = bd_tank$distances
) %>%
  mutate(tank.x = as.character(tank.x)) %>%            # force character
  mutate(tank.x = reorder(tank.x, Distance, FUN = median))  # reorder for plotting

# 2) Median labels for just NWQ4 and P4
med_df <- disp_all %>%
  mutate(tank_chr = as.character(tank.x)) %>%          # stable tank name
  filter(tank_chr %in% c("NWQ4", "P4")) %>%
  dplyr::group_by(tank_chr) %>%
  dplyr::summarise(
    med = median(Distance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y_lab = med + 0.001,                               # nudge up; adjust if needed
    lab   = sprintf("median = %.3f", med)
  )

# 3) Fill colors: grey for all tanks, with NWQ4 red and P4 blue
tank_levels <- levels(disp_all$tank.x)
fill_vals <- setNames(rep("grey80", length(tank_levels)), tank_levels)
fill_vals["NWQ4"] <- "#d73027"
fill_vals["P4"]   <- "#4575b4"

annot_text <- "betadisper by tank: F = 3.16, p = 0.0022"

# 4) Plot
ggplot(disp_all, aes(x = tank.x, y = Distance, fill = tank.x)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.6) +
  scale_fill_manual(values = fill_vals, guide = "none") +
  geom_text(
    data = med_df,
    aes(x = tank_chr, y = y_lab, label = lab),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.5
  ) +
  theme_bw() +
  labs(
    x = "Tank",
    y = "Distance to centroid",
    title = "Within-tank microbiome variability"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = annot_text,
    hjust = 1.05, vjust = 1.5,
    size = 3.5
  )
#ggsave(filename = "outputs/acer_boxplot_ALL.pdf", plot = last_plot())

#plot same as above with geno points color ####
# 1) Per-sample dispersion table
disp_all <- tibble(
  Sample   = rownames(Bac.seq.rel.wu.df),
  tank.x   = Bac.seq.rel.wu.df$tank.x,
  Geno     = Bac.seq.rel.wu.df$Geno,
  Distance = bd_tank$distances
) %>%
  mutate(
    tank.x = as.character(tank.x),
    tank.x = reorder(tank.x, Distance, FUN = median)
  )

# 2) Median labels for NWQ4 and P4
med_df <- disp_all %>%
  filter(tank.x %in% c("NWQ4", "P4")) %>%
  dplyr::group_by(tank.x) %>%
  dplyr::summarise(
    med = median(Distance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y_lab = med + 0.001,
    lab   = sprintf("median = %.3f", med)
  )

# 3) Box fill colors (tank-level)
tank_levels <- levels(factor(disp_all$tank.x))
fill_vals <- setNames(rep("grey80", length(tank_levels)), tank_levels)
fill_vals["NWQ4"] <- "#d73027"
fill_vals["P4"]   <- "#4575b4"

annot_text <- "betadisper by tank: F = 3.16, p = 0.0022"

# 4) Plot
ggplot(disp_all, aes(x = tank.x, y = Distance)) +
  # boxplots (tank-level)
  geom_boxplot(
    aes(fill = tank.x),
    outlier.shape = NA,
    alpha = 0.7
  ) +
  # points colored by genotype
  geom_jitter(
    aes(color = Geno),
    width = 0.15,
    size = 1.8,
    alpha = 0.7
  ) +
  scale_fill_manual(values = fill_vals, guide = "none") +
  theme_bw() +
  labs(
    x = "Tank",
    y = "Distance to centroid",
    title = "Within-tank microbiome variability",
    color = "Genotype"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # median labels
  geom_text(
    data = med_df,
    aes(x = tank.x, y = y_lab, label = lab),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.5
  ) +
  # annotation
  annotate(
    "text",
    x = Inf, y = Inf,
    label = annot_text,
    hjust = 1.05, vjust = 1.5,
    size = 3.5
  )

#ggsave(filename = "outputs/acer_boxplot_GEno.pdf", plot = last_plot())

#plot same as above with DATE points color ####

# 1) Build per-sample dispersion table from betadisper output + attach metadata needed for coloring
disp_all <- data.frame(
  Sample   = rownames(Bac.seq.rel.wu.df),
  tank.x   = Bac.seq.rel.wu.df$tank.x,
  Date     = Bac.seq.rel.wu.df$Date,
  Distance = bd_tank$distances
) %>%
  mutate(
    Date   = as.Date(Date),  # safe even if already Date
    Date_f = factor(Date, levels = sort(unique(Date))),
    tank.x = reorder(tank.x, Distance, FUN = median)
  )

# 2) Median labels for just NWQ4 and P4
med_df <- disp_all %>%
  filter(as.character(tank.x) %in% c("NWQ4", "P4")) %>%
  dplyr::group_by(tank.x) %>%
  dplyr::summarise(med = median(Distance, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    y_lab = med + 0.001,                 # nudge up; adjust if needed
    lab   = sprintf("median = %.3f", med)
  )

# 3) Tank fill colors (boxes): grey for all tanks, NWQ4 red, P4 blue
tank_levels <- levels(disp_all$tank.x)
fill_vals <- setNames(rep("grey80", length(tank_levels)), tank_levels)
fill_vals["NWQ4"] <- "#d73027"
fill_vals["P4"]   <- "#4575b4"

# 4) Date colors (points): rainbow in chronological order
date_levels <- levels(disp_all$Date_f)
date_cols <- setNames(grDevices::rainbow(length(date_levels)), date_levels)

annot_text <- "betadisper by tank: F = 3.16, p = 0.0022"

# ensure Date is a factor in chronological order
disp_all <- disp_all %>%
  mutate(Date_f = factor(Date, levels = sort(unique(Date))))

ggplot(disp_all, aes(x = tank.x, y = Distance)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, fill = "grey85") +
  geom_jitter(
    aes(color = Date_f),
    width = 0.15,
    size = 1.8,
    alpha = 0.8
  ) +
  scale_color_viridis_d(option = "plasma", end = 0.95) +
  theme_bw() +
  labs(
    x = "Tank",
    y = "Distance to centroid",
    title = "Within-tank microbiome variability",
    color = "Sampling date"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#ggsave(filename = "outputs/acer_boxplot_Date.pdf", plot = last_plot())


```

#Alpha Diversity
```{r}
#Shannon, richness, evenness (Pielou’s J)
#using counts data Bac.seq.ACER
ps <- Bac.seq.ACER   # change this to your counts object

# Richness + Shannon from phyloseq
alpha_df <- estimate_richness(ps, measures = c("Observed", "Shannon")) %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::rename(
    Richness = Observed,
    Shannon  = Shannon
  ) %>%
  dplyr::mutate(
    Evenness = ifelse(Richness > 0, Shannon / log(Richness), NA_real_)
  )

alpha_df <- alpha_df %>%
  dplyr::mutate(
    SampleID = sub("^X", "", SampleID)
  )
# Add metadata
meta_df <- as(sample_data(ps), "data.frame") %>%
  tibble::rownames_to_column("SampleID")

alpha_df <- alpha_df %>%
  left_join(meta_df, by = "SampleID")

alpha_df

#Models ####
# Shannon
hist(alpha_df$Shannon)
hist(log(alpha_df$Shannon)) #good

alpha_df <- alpha_df %>%
  mutate(
    logShannon = log(Shannon),
    Date = as.factor(Date),
    tank.x = as.factor(tank.x),
    Geno = as.factor(Geno)
  )

m2 <- lmer(logShannon ~ tank.x *Date + (1 | Geno), data = alpha_df)
anova(m2)

par(mfrow = c(2,2)) #check residuals
plot(m2) # good


# Richness
hist(alpha_df$Richness)
hist(log(alpha_df$Richness)) #good

alpha_df <- alpha_df %>%
  mutate(
    logRichness = log(Richness),
    Date = as.factor(Date),
    tank.x = as.factor(tank.x),
    Geno = as.factor(Geno)
  )

m2 <- lmer(logRichness ~ tank.x *Date + (1 | Geno), data = alpha_df)
anova(m2)

par(mfrow = c(2,2)) #check residuals
plot(m2) # this is ok, but not great...lets try to deal with this using GLMM ->

library(glmmTMB)
library(car)
m_rich_nb <- glmmTMB(
  Richness ~ tank.x * Date + (1 | Geno),
  data = alpha_df,
  family = nbinom2()
)

Anova(m_rich_nb, type = 2)


# Evenness
hist(alpha_df$Evenness)
hist(log(alpha_df$Evenness)) #good


alpha_df <- alpha_df %>%
  mutate(logEvenness = log(Evenness))

options(contrasts = c("contr.sum", "contr.poly"))
m_ev
en <- lmer(
  logEvenness ~ tank.x * Date + (1 | Geno),
  data = alpha_df
)

anova(m_even) #interaction not sig



```
Alpha Plots
```{r}
#plots ####
# make sure Date is an ordered factor
alpha_df <- alpha_df %>%
  dplyr::mutate(
    Date_f = factor(Date, levels = sort(unique(Date)))
  )

# Reuse the SAME Date levels
date_levels <- levels(alpha_df$Date_f)

scale_color_viridis_d(option = "plasma", end = 0.95)


# Use the SAME palette you used before (example: rainbow)
date_pal <- setNames(
  grDevices::rainbow(length(date_levels)),
  date_levels
)

library(viridis)

#Richness ####
ggplot(alpha_df, aes(x = tank.x, y = Richness)) +
  # boxplots by Date within tank
  geom_boxplot(
    aes(fill = Date_f, group = interaction(tank.x, Date_f)),
    position = position_dodge(width = 0.8),
    outlier.shape = NA,
    alpha = 0.5
  ) +

  # points overlaid
  geom_jitter(
    aes(color = Date_f),
    position = position_jitterdodge(
      jitter.width = 0.15,
      dodge.width  = 0.8
    ),
    size = 2,
    alpha = 0.8
  ) +

  # viridis colors (same as your other plots)
  scale_fill_viridis_d(option = "D", end = 0.95) +
  scale_color_viridis_d(option = "D", end = 0.95) +

  theme_bw() +
  labs(
    x = "Tank",
    y = "Richness",
    fill  = "Date",
    color = "Date"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(filename = "outputs/acer_richness.pdf",plot = last_plot(),width = 10,height = 5,units = "in")

#Evenness
ggplot(alpha_df, aes(x = tank.x, y = Evenness)) +
  # boxplots by Date within tank
  geom_boxplot(
    aes(fill = Date_f, group = interaction(tank.x, Date_f)),
    position = position_dodge(width = 0.8),
    outlier.shape = NA,
    alpha = 0.5
  ) +

  # points overlaid
  geom_jitter(
    aes(color = Date_f),
    position = position_jitterdodge(
      jitter.width = 0.15,
      dodge.width  = 0.8
    ),
    size = 2,
    alpha = 0.8
  ) +

  # viridis colors (same as your other plots)
  scale_fill_viridis_d(option = "D", end = 0.95) +
  scale_color_viridis_d(option = "D", end = 0.95) +

  theme_bw() +
  labs(
    x = "Tank",
    y = "Evenness",
    fill  = "Date",
    color = "Date"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave(filename = "outputs/acer_evennenss.pdf",plot = last_plot(),width = 10,height = 5,units = "in")

#Shannnon
ggplot(alpha_df, aes(x = tank.x, y = Shannon)) +
  # boxplots by Date within tank
  geom_boxplot(
    aes(fill = Date_f, group = interaction(tank.x, Date_f)),
    position = position_dodge(width = 0.8),
    outlier.shape = NA,
    alpha = 0.5
  ) +

  # points overlaid
  geom_jitter(
    aes(color = Date_f),
    position = position_jitterdodge(
      jitter.width = 0.15,
      dodge.width  = 0.8
    ),
    size = 2,
    alpha = 0.8
  ) +

  # viridis colors (same as your other plots)
  scale_fill_viridis_d(option = "D", end = 0.95) +
  scale_color_viridis_d(option = "D", end = 0.95) +

  theme_bw() +
  labs(
    x = "Tank",
    y = "Shannon",
    fill  = "Date",
    color = "Date"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

#ggsave(filename = "outputs/acer_shannon.pdf",plot = last_plot(),width = 10,height = 5,units = "in")


```
# ANCOM
ANCOM-BC2
Which taxa differ in abundance between tanks (esp NWQ4), after accounting for Date and Geno?

Running into a lot of issue with this code. come back to it. 
```{r}
library(ANCOMBC)
ps_counts<-Bac.seq.ACER #dup
sample_data(ps_counts)$Date <- as.Date(sample_data(ps_counts)$Date, format = "%m/%d/%y")
sample_data(ps_counts)$Date <- factor(sample_data(ps_counts)$Date)
sample_data(ps_counts)$tank.x <- factor(sample_data(ps_counts)$tank.x)
sample_data(ps_counts)$Geno <- factor(sample_data(ps_counts)$Geno)

sd <- data.frame(sample_data(ps_counts)) #to compare T0 to everything else laster
sd$Date <- as.Date(sd$Date, format = "%m/%d/%y")
sd$TimePhase <- ifelse(sd$Date == as.Date("2023-04-03"), "T0", "Post")
sd$TimePhase <- factor(sd$TimePhase, levels = c("T0","Post"))
sample_data(ps_counts) <- sample_data(sd)

#ANCOM-BC2, Full model, tank.x+Date+interaction+genotype (covariate)
library(phyloseq)
library(dplyr)
library(ANCOMBC)

# Grab sample_data as a plain data.frame
sd <- data.frame(sample_data(ps_counts))
sd$SampleID <- rownames(sd)

# Keep only samples that are in ps_counts (should already be true, but safe)
sd <- sd %>%
  filter(SampleID %in% sample_names(ps_counts)) %>%
  filter(!is.na(tank.x) & tank.x != "") %>%
  mutate(
    tank.x = droplevels(factor(tank.x)),
    Geno   = droplevels(factor(Geno))
  )

# Prune phyloseq to the same samples and re-attach cleaned metadata
ps_counts2 <- prune_samples(sd$SampleID, ps_counts)
sample_data(ps_counts2) <- sample_data(sd %>% column_to_rownames("SampleID"))

#  checks
table(sample_data(ps_counts2)$tank.x)
stopifnot(all(sample_names(ps_counts2) == rownames(sample_data(ps_counts2))))

set.seed(30)

res_tank_global <- ancombc2(
  data = ps_counts2,
  assay_name = "otu_table",
  tax_level = NULL,
  fix_formula = "Geno",
  group = "tank.x",
  rand_formula = NULL,
  p_adj_method = "BH",
  prv_cut = 0.05,
  lib_cut = 3000,
  struc_zero = TRUE,
  neg_lb = TRUE,
  global = TRUE,
  pairwise = FALSE,   # start here
  dunnet = FALSE,
  trend = FALSE
)

set.seed(30)

library(ANCOMBC)

res_tank_global <- ancombc2(
  data = ps_counts2,
  assay_name = "otu_table",
  tax_level = NULL,
  fix_formula = "Geno",
  group = "tank.x",
  p_adj_method = "BH",
  prv_cut = 0.01,
  lib_cut = 3000,
  struc_zero = FALSE,   # keep off for stability
  neg_lb = TRUE,
  global = TRUE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE
)

res_tank_global$res_global |> 
  dplyr::filter(diff_robust_abn == TRUE) |> 
  dplyr::arrange(q_val)



```

#run linear models on individual taxa of interest:
```{r}
Bac.seq.rel2<-Bac.seq.rel

sd <- as(sample_data(Bac.seq.rel2), "data.frame")  # plain DF
sd <- as.data.frame(sd)                             # extra-safe

sd <- sd %>%
  mutate(
    Date = mdy(Date),  # parses "4/3/23"
    DayNum = as.integer(Date - min(Date, na.rm = TRUE))
  )

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

otu_long <- psmelt(ps_rel) %>%
  as_tibble() %>%
  rename(
    SampleID = Sample,
    Taxon = OTU,
    RelAbund = Abundance
  ) %>%
  left_join(sd, by = "SampleID")

library(tibble)
library(lubridate)
library(dplyr)
library(tidyr)
library(purrr)
library(lme4)
library(lmerTest)

sd <- as(sample_data(Bac.seq.rel2), "data.frame") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("SampleID") %>%
  mutate(
    Date = mdy(Date),
    DayNum = as.integer(Date - min(Date, na.rm = TRUE)),
    tank.x = factor(tank.x),
    Geno   = factor(Geno)
  )


############# start again

library(phyloseq)
library(dplyr)
library(tibble)
library(lubridate)

# metadata as plain DF
sd <- as(sample_data(Bac.seq.rel2), "data.frame") |> 
  as.data.frame() |>
  mutate(
    Date = mdy(Date),  # change to ymd() if your strings are "2023-04-03"
    DayNum = as.integer(Date - min(Date, na.rm = TRUE))
  ) |>
  tibble::rownames_to_column("SampleID")  # only if rownames are sample IDs

otu_long <- psmelt(Bac.seq.rel2) |>
  as_tibble() |>
  rename(
    SampleID = Sample_ID,
    Taxon    = OTU,
    RelAbund = Abundance
  ) |>
  left_join(sd, by = "SampleID")

#top 10
top10 <- otu_long |>
  group_by(Taxon) |>
  summarise(mean_rel = mean(RelAbund, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(mean_rel)) |>
  slice_head(n = 10) |>
  pull(Taxon)

otu_top <- otu_long |>
  filter(Taxon %in% top10) |>
  mutate(
    tank.x = factor(tank.x.y),
    Geno   = factor(Geno.x)
  )

#models ####
library(lmerTest)

eps <- 1e-6

otu_top <- otu_top |>
  mutate(
    RelAbund_clip = pmin(pmax(RelAbund, eps), 1 - eps),
    logitRel      = qlogis(RelAbund_clip)
  )

mods <- otu_top |>
  group_by(Taxon) |>
  group_map(~{
    m <- lmer(logitRel ~ tank.x * DayNum + (1 | Geno), data = .x)
    out <- anova(m)
    list(model = m, anova = out)
  })
names(mods) <- unique(otu_top$Taxon)

library(lmerTest)

anova_tbl <- otu_top %>%
  group_by(Taxon) %>%
  group_modify(~{
    m <- lmer(logitRel ~ tank.x * DayNum + (1 | Geno), data = .x)
    out <- as.data.frame(anova(m))
    out$term <- rownames(out)
    rownames(out) <- NULL
    out
  }) %>%
  ungroup()

#write.csv(anova_tbl, "anova_tbl.csv")

#models per OTU that was significant for tank (OTUS p<0.05) ####
sig_taxa <- c("Otu0013", "Otu0031", "Otu0124")
otu_sig <- otu_top |>
  filter(Taxon %in% sig_taxa)

taxon_posthoc <- otu_sig |>
  group_by(Taxon) |>
  nest() |>
  mutate(
    model = map(data, ~
      lmer(
        logitRel ~ tank.x.y * DayNum + (1 | Geno.x),
        data = .x
      )
    ),

    # marginal tank means
    emm_tank = map(model, ~ emmeans(.x, ~ tank.x.y)),

    # Tukey-adjusted pairwise tank contrasts
    tank_pairs = map(
      emm_tank,
      ~ pairs(.x, adjust = "tukey") |> as.data.frame()
    )
  )

tank_contrasts <- taxon_posthoc |>
  select(Taxon, tank_pairs) |>
  unnest(tank_pairs)

#write.csv(tank_contrasts, "tank_contrasts.csv")

```
Plots of individual top 10 taxa
```{r}
# 1) Make sure Date is a real Date
otu_top2 <- otu_top %>%
  mutate(
    Date = as.Date(Date.x),                    # if already Date
    # If Date is like "4/3/23", use this instead:
    # Date = mdy(Date),
    tank.x = factor(tank.x)
  )

# 2) Summarize mean +/- SE per Taxon x Tank x Date
sum_top <- otu_top2 %>%
  group_by(Taxon, tank.x.x, Date.x) %>%
  summarise(
    mean_rel = mean(RelAbund, na.rm = TRUE),
    se_rel   = sd(RelAbund, na.rm = TRUE) / sqrt(sum(!is.na(RelAbund))),
    n        = sum(!is.na(RelAbund)),
    .groups  = "drop"
  )

# Optional: order tanks by overall mean within each taxon (helps readability)
# If you want a single consistent tank order across plots, comment this out.
tank_order <- otu_top2 %>%
  dplyr::group_by(tank.x.x) %>%
   dplyr::summarise(overall_mean = mean(RelAbund, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(overall_mean)) %>%
  pull(tank.x.x)
sum_top <- sum_top %>% mutate(tank.x = factor(tank.x.x, levels = tank_order))

# 3A) One figure, faceted (each taxon its own panel)
p_facet <- ggplot(sum_top, aes(x = Date.x, y = mean_rel, group = tank.x, color = tank.x)) +
  geom_line(linewidth = 0.8, alpha = 0.9) +
  geom_point(size = 2, alpha = 0.9) +
  geom_ribbon(
    aes(ymin = pmax(mean_rel - se_rel, 0), ymax = pmin(mean_rel + se_rel, 1), fill = tank.x),
    alpha = 0.15,
    color = NA
  ) +
  facet_wrap(~ Taxon, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Date",
    y = "Mean relative abundance",
    color = "Tank",
    fill  = "Tank",
    title = "Top 10 taxa trajectories by tank over time"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_facet

# Save the faceted figure wider
ggsave("outputs/top10_taxa_tank_time_faceted.pdf", p_facet, width = 14, height = 9)

# 3B) One PDF per taxon (each taxon gets its own plot)
dir.create("outputs/top10_taxa_by_tank", recursive = TRUE, showWarnings = FALSE)

taxa_list <- unique(sum_top$Taxon)

for (tx in taxa_list) {
  p_one <- ggplot(filter(sum_top, Taxon == tx),
                  aes(x = Date, y = mean_rel, group = tank.x, color = tank.x)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    geom_ribbon(
      aes(ymin = pmax(mean_rel - se_rel, 0), ymax = pmin(mean_rel + se_rel, 1), fill = tank.x),
      alpha = 0.15,
      color = NA
    ) +
    theme_bw() +
    labs(
      x = "Date",
      y = "Mean relative abundance",
      color = "Tank",
      fill  = "Tank",
      title = paste("Taxon:", tx)
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggsave(
    filename = file.path("outputs/top10_taxa_by_tank", paste0(tx, "_tank_time.pdf")),
    plot = p_one,
    width = 10, height = 6
  )
}

# try again wtih Date spaced out####

date_levels <- sort(unique(sum_top$Date.x))

date_levels_spaced <- c(
  as.character(date_levels[-length(date_levels)]),
  "gap",
  as.character(date_levels[length(date_levels)])
)

sum_top <- sum_top %>%
  mutate(
    Date_f = factor(
      as.character(Date.x),
      levels = date_levels_spaced,
      ordered = TRUE
    )
  )

###########
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

# --- 0) Make sure Date.x is a real Date and RelAbund is numeric ---
otu_long2 <- otu_long %>%
  mutate(
    Date.x = as.Date(Date.x),            # if this fails, swap to mdy(Date.x)
    RelAbund = as.numeric(RelAbund)
  )

# Check quickly (should be 0..1-ish, not near 1 for a single OTU)
summary(otu_long2$RelAbund)

# --- 1) Top 10 taxa by mean relative abundance ---
top10 <- otu_long2 %>%
  group_by(Taxon) %>%
  summarise(mean_rel = mean(RelAbund, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rel)) %>%
  slice_head(n = 10) %>%
  pull(Taxon)

otu_top <- otu_long2 %>%
  filter(Taxon %in% top10) %>%
  dplyr::mutate(
    tank.x.x = factor(tank.x.x)
  )

# --- 2) Summarize to one value per tank per date (mean ± SE) ---
sum_top <- otu_top %>%
  group_by(Taxon, tank.x.x, Date.x) %>%
  summarise(
    mean_rel = mean(RelAbund, na.rm = TRUE),
    se_rel   = sd(RelAbund, na.rm = TRUE) / sqrt(sum(!is.na(RelAbund))),
    .groups = "drop"
  )

# --- 3) Build DatePlot with extra spacing BEFORE the last date ---
date_levels <- sort(unique(sum_top$Date.x))
last_date   <- max(date_levels)

sum_top <- sum_top %>%
  mutate(
    DatePlot = if_else(Date.x == last_date, as.character(Date.x), as.character(Date.x)),
    DatePlot = factor(DatePlot, levels = c(as.character(date_levels[date_levels != last_date]),
                                          " ",  # blank spacer
                                          as.character(last_date)))
  )

# --- 4) Plot (one panel per taxon, colored by tank) ---
p <- ggplot(sum_top, aes(x = DatePlot, y = mean_rel, color = tank.x.x, group = tank.x.x)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  geom_errorbar(aes(ymin = mean_rel - se_rel, ymax = mean_rel + se_rel), width = 0.1, alpha = 0.6) +
  facet_wrap(~ Taxon, scales = "free_y", ncol = 4) +
  theme_bw() +
  labs(
    x = "Date",
    y = "Mean relative abundance (± SE)",
    color = "Tank",
    title = "Top 10 taxa trajectories by tank",
    subtitle = "Extra spacing inserted before the last sampling date"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  );p
#ggsave(filename = "outputs/acer_topten_lines.pdf", plot = last_plot())





#plot of JUST OTU0001


# 0) Make sure Date.x is character (raw strings like "4/3/23")
otu_top <- otu_top %>%
  mutate(
    Date_raw = as.character(Date.x),
    Date_raw = trimws(Date_raw)
  )

# 1) Parse using the format you actually have: m/d/yy
otu_top <- otu_top %>%
  mutate(
    Date_real = as.Date(Date_raw, format = "%m/%d/%y")
  )

# 2) Sanity checks
if (any(is.na(otu_top$Date_real))) {
  bad <- unique(otu_top$Date_raw[is.na(otu_top$Date_real)])
  stop("Some Date.x values did not parse. Examples: ", paste(bad, collapse = ", "))
}

# 3) Optional: lock the date ordering you want (chronological)
date_levels <- as.Date(c("4/3/23","4/5/23","4/7/23","4/11/23","5/15/23"), format = "%m/%d/%y")
otu_top <- otu_top %>%
  mutate(
    Date_f = factor(Date_real, levels = date_levels)
  )

fig1 <- sum_top %>%
  filter(Taxon == "Otu0001") %>%
  ggplot(aes(
    x = Date.x,
    y = mean_rel,
    color = tank.x,
    group = tank.x
  )) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_date(
    date_breaks = "1 week",
    date_labels = "%b %d"
  ) +
  labs(
    x = "Sampling date",
    y = "Relative abundance",
    title = "Dominant OTU (Otu0001) across tanks"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank()
  )

fig1


```



